### 2017-01-27 README.Rmd
Finalize vignette / README

This package allows you to explore and map data on county-level exposures to Atlantic-basin tropical storms between 1988 and 2015 for a number of storm hazards (e.g., wind, rain, flood events, distance from the storm track) for counties in the eastern half of the United States. This package allows the user to map exposures by county for a selected storm or to identify all storms to which selected counties were exposed, based on user-specified thresholds (for example, the package allows the user to identify all storms that brought wind of 34 knots or higher to Miami-Dade County in Florida).

@@

This package depends on data in a data package (`hurricaneexposuredata`) that is available through a `drat` repository on GitHub. To use the `hurricaneexposure` package, you will need to install `hurricaneexposuredata` on your computer. You can do that by adding the `drat` repository to your repositories; once you do this, you can install the `hurricaneexposuredata` using the `install.packages` function (and later update it using the `update.packages` function):

@@

- `county_centers`: Location of United States county centers of population
- `hurr_tracks`: Storm tracks for Atlantic-basin storms, 1988-2015
- `closest_dist`: Closest distances between counties and a storm track, for Atlantic-basin storms, 1988-2015
- `rain`: Rainfall for US counties during Atlantic basin tropical storms, 1988-2011; daily rainfall is given from five days before to three days after the storm's closest approach to the county
- `storm_winds`: Modeled county wind speeds for Atlantic-basin storms, 1988-2015
- `storm_events`: Listings from the NOAA Storm Events database that occurred near in time and location to tropical storms, 1988-2015. This database changed the types of events it reported in 1996, which should be considered when using the data. 
- `ext_tracks_wind`: Estimated county wind speeds for Atlantic-basin storms, 1988-2015, based on the wind radii listed in the Extended Best Tracks dataset

Once you've installed and loaded `hurricaneexposuredata`, you can load the included data using the `data` function. For example:

@@

For each dataset included in `hurricaneexposuredata`, you can see the helpfiles for the data for more information about the data included in each (e.g., `?hurr_tracks` to read documentation on the hurricane tracks data shown above). This data documentation includes both explanations of how the dataset was created, definitions of the variables included in each dataset, and the sources of data included.

@@

In functions throughout the `hurricaneexposure` package, storms are identified based on their name and year (e.g., Hurricane Floyd in 1999 is identified as "Floyd-1999"). Note that a few storms satisfied this criteria but were unnamed (e.g., "Notnamed" in 1991, "Subtrop" in 1997, etc.). These storms are identified in functions in the `hurricaneexposure` package based on the identifying name listed in the above table (e.g., "Notnamed-1991"), although for many applications you may want to consider excluding these unnamed storms from the full analysis.

@@

The `hurricaneexposure` package has functions to interact with the data in the `hurricaneexposuredata` package. First, this package has functions that can be used to create some different maps of hurricane exposures based on distance to the storm track and rainfall.

@@

First, for several of the storm hazard measurements (currently rain, wind, and distance), you can use the `map_counties` function in `hurricaneexposure` to plot a map for a specific storm showing the continuous value of that metric for all eastern U.S. counties. For example, you can map county-level rain during Hurricane Floyd (1999) using the following call:

@@

By default, this map shows the cumulative rain in each county for two days before to one day after the date that the storm passed closest to the county (see the help file for the `closest_dist` dataset in `hurricaneexposuredata` to find out more about how this date was determined for each county-storm combination). The period used to calculate cumulative rainfall for the map can be customized using the `days_included` option in the `map_counties` function. The two following examples show the difference in rain maps for Tropical Storm Allison (2001) when using rain only from the day before and the day the storm was closest to each county (`days_included = -1:0`) versus from five days before to three days after the storm's closest approach (`days_included = -5:3`):

@@

These county winds are determined based on a model of wind speeds from Best Tracks hurricane data. See the documentation for the `storm_winds` dataset (`?storm_winds`) to find out more about this data; further details on the wind model are available in the vignettes for the [`stormwindmodel` package](https://cran.r-project.org/web/packages/stormwindmodel/index.html).

[Add details about how to map this from the wind radii estimates, instead]

Finally, you can also use this function to plot the closest distance between the storm and each county. For this, you use the argument `metric = "distance"`. For example, the following call plots county-level distances from the track of Hurricane Sandy (2012):

@@

For example, you can map a binary variable of distance-based exposure using `map_distance_exposure`, with "exposure" defined as the county being within the number of kilometers given in `dist_limit` of the storm track. For example, you can map the counties that were within 75 km of Hurricane Sandy's storm track with the call:

@@

Similarly, you can map a binary variable county exposure based on rain using `map_rain_exposure`. In the case of rain, you must define exposure thresholds for both rainfall and distance. This is because the rain data is observed data, so neglecting to limit "exposed" counties to those close to the storm track may pick up counties far from the storm that experienced a lot of rain over the same period. For example, you can use the following call to map rain exposure to Tropical Storm Allison (2001), where exposure is defined as a county where cumulative rainfall from fives days before to three days after the storm's closest approach was 175 mm or more and where the county was within 500 km of the storm's track:

@@

You can use the `map_wind_exposure` function to map counties exposed to a certain level of wind during a storm. For example, to identify all counties exposed to sustained winds of 20 m / s or higher during Hurricane Katrina, you can run:

@@

Event listings from the NOAA Storm Events database were linked to storms based on the start date of the event being within a five-day window of the date the storm was closest to the county and the county being within 500 km of the county. The [`noaastormevents` package](https://github.com/zailchen/noaastormevents) was used to create this dataset. For more information on the data, see the help file for the `storm_events` dataset (`?storm_events`) or the vignette for the `noaastormevents` package.

@@

If you would like to add a storm's track to one of the maps of county-level exposure, you can do so using the `map_tracks` function. This function can be used either to map a storm track on a blank U.S. map or to add a track to an existing map, if it is a `ggplot` object (as are all maps generated in the previous section).

@@

The `map_tracks` function is used in all the mapping functions described in the previous function to add the storm's track to the exposure map. However, if you would like to customize the appearance of the storm's track on the map, you can do so by plotting the exposure map without the storm track (`add_track = FALSE`) and then adding the track yourself with the `map_tracks` function. For example, to change the color of the storm track in a map of flood event exposure during Hurricane Floyd (1999) and to add points showing the available observations for the storm in the Best Tracks hurricane tracking data, you can run:

@@

The `hurricaneexposure` package also has several functions that can input a list of counties and output a list of all of the storms each county was exposed to over a certain period, based on one of the hazard metrics. These exposure listings include the storm dates, making them easy to integrate into time series datasets of health or other impact data.

For example, the `county_rain` function takes a list of county FIPS codes, bounds on the starting and ending years of the analysis, and thresholds to define rain-based exposure (cumulative rainfall and distance from the storm's track, as explained for the binary rain maps shown in an earlier section) and creates a list of all storms that met these thesholds for the counties. For example, to get a dataset of all the storms to which Orleans Parish (FIPS 22071), and Newport News, Virginia (FIPS 51700), were exposed between 1995 and 2005, where "exposed" means that the storm passed within 100 kilometers of the county center and the rainfall over a three-day window of the date of closest approach was 100 millimeters or more, you can run:

@@

The returned dataframe includes the estimated maximum values during the storm for sustained winds (`vmax_sust`) and gust winds (`vmax_gust`), both in m / s. It also includes some of the variables on the date and time of closest approach and the distance at the storm's closest approach.

@@

Similar functions exist for wind exposure (`multi_county_wind`) and distance exposure (`multi_county_distance`).

>>>>

This package allows users to explore and map data of county-level exposures to Atlantic-basin tropical storms between 1988 and 2015 for a number of storm hazards (e.g., wind, rain, flood events, distance from the storm track) for counties in the eastern half of the United States. Users can map exposures by county for a selected storm and can also identify all storms to which selected counties were exposed, based on user-specified thresholds (for example, the package allows the user to identify all storms that brought wind of 34 knots or higher to Miami-Dade County in Florida).

@@

This package depends on data in a data package (`hurricaneexposuredata`) that is available through a `drat` repository on GitHub. To use the `hurricaneexposure` package, you will need to install `hurricaneexposuredata` on your computer. You can do that by adding the `drat` archive to the list of repositories your system will query when adding and updating R packages. Once you do this, you can install the `hurricaneexposuredata` using the `install.packages` function (and later update it using the `update.packages` function):

@@

- `county_centers`: Location of United States county centers of population
- `hurr_tracks`: Storm tracks for Atlantic-basin storms, 1988-2015
- `closest_dist`: Closest distances between counties and a storm track, for Atlantic-basin storms, 1988-2015
- `rain`: Rainfall for US counties during Atlantic basin tropical storms, 1988-2011; daily rainfall is given from five days before to three days after the storm's closest approach to the county
- `storm_winds`: Modeled county wind speeds for Atlantic-basin storms, 1988-2015
- `storm_events`: Listings from the [NOAA Storm Events database](https://www.ncdc.noaa.gov/stormevents/) that occurred near in time and location to tropical storms, 1988-2015. This database changed the types of events it reported in 1996, which should be considered when using the data. 
- `ext_tracks_wind`: Estimated county wind speeds for Atlantic-basin storms, 1988-2015, based on the wind radii listed in the [Extended Best Tracks dataset](http://rammb.cira.colostate.edu/research/tropical_cyclones/tc_extended_best_track_dataset/)

Once you've installed and loaded `hurricaneexposuredata`, you can load the included data using the `data` function. For example, you can access the data on hurricane tracks (`hurr_tracks`) using:

@@

For each dataset included in `hurricaneexposuredata`, you can see the helpfiles for the data for more information (e.g., `?hurr_tracks` to read documentation on the hurricane tracks data shown above). This data documentation includes both explanations of how the dataset was created, definitions of the variables included in each dataset, and the sources of data included.

@@

In functions throughout the `hurricaneexposure` package, storms are identified based on their name and year (e.g., Hurricane Floyd in 1999 is identified as "Floyd-1999"). Note that a few storms satisfied the criteria to be included in the data but were unnamed (e.g., "Notnamed" in 1991, "Subtrop" in 1997, etc.). These storms are identified in functions in the `hurricaneexposure` package based on the identifying name listed in the above table (e.g., "Notnamed-1991"), although for some research projects you may want to consider excluding these unnamed storms from the analysis.

@@

The `hurricaneexposure` package has functions to interact with the data in the `hurricaneexposuredata` package. First, there are functions that can be used to create some different maps of hurricane exposures based on distance to the storm track and rainfall.

@@

For several of the storm hazard measurements (currently rain, wind, and distance), you can use the `map_counties` function in `hurricaneexposure` to plot a map for a specific storm showing the continuous value of that metric for all eastern U.S. counties. For example, you can map county-level rain during Hurricane Floyd (1999) using the following call:

@@

By default, this map shows the cumulative rain in each county for two days before to one day after the date that the storm passed closest to the county (see the help file for the `closest_dist` dataset in `hurricaneexposuredata` to find out more about how this date of closest approach was determined for each county-storm combination). The period used to calculate cumulative rainfall for the map can be customized using the `days_included` option in the `map_counties` function. The two following examples show the difference in rain maps for Tropical Storm Allison (2001) when using rain only from the day before and the day the storm was closest to each county (`days_included = -1:0`) versus from five days before to three days after the storm's closest approach (`days_included = -5:3`):

@@

These county winds are determined based on a model of wind speeds, using as input data from the hurricane tracks data. See the documentation for the `storm_winds` dataset (`?storm_winds`) to find out more about this data; further details on the wind model are available in the vignettes for the [`stormwindmodel` package](https://cran.r-project.org/web/packages/stormwindmodel/index.html).

You can map a few other variables of wind with this function. For example, you can map the duration of winds of 20 m / s or more by using the argument `wind_var = "sust_dur"`:

```{r fig.width = 7, fig.height = 4}
map_counties("Katrina-2005", metric = "wind", wind_var = "sust_dur")
```

You can map estimated gust winds, rather than sustained winds (the default), using the argument `wind_var = "vmax_gust"`. Further, we have included a second source of estimated winds in the data available in `hurricaneexposuredata`. These wind estimates are based on the wind radii in the Extended Best Tracks dataset (you can find out more about this data in the helpfile for the `ext_tracks_wind` dataset). These data provide estimates of which counties were exposured during a storm to sustained winds in four categories: 0--34 knots; 34--50 knots;  50--64 knots; and 64 knots or higher. Therefore, these data provide a categorical rather than continuous estimate of county wind speeds. However, they may be preferable to the model wind speeds for some storms, especially storms in extra-tropical transition. Further, you may find it interesting to use these data in a sensitivity analysis, to compare if impact study results are sensitive to whether these wind data or the modeled wind data are used for exposure classification. You can create a wind map based on this data using the `wind_source = "ext_tracks"` argument in the `map_counties` function. For example, to map estimated winds during Hurricane Katrina based on the Extended Best Tracks wind radii, you can run:

```{r fig.width = 7, fig.height = 4}
map_counties("Katrina-2005", metric = "wind", wind_source = "ext_tracks")
```

Finally, you can also use the `map_counties` function to plot the closest distance between the storm and each county. For this, you use the argument `metric = "distance"`. For example, the following call plots county-level distances from the track of Hurricane Sandy (2012):

@@

For example, you can map a binary variable of distance-based exposure using `map_distance_exposure`, with "exposure" defined as that the county was within the number of kilometers given in `dist_limit` of the storm track. For example, you can map the counties that were within 75 km of Hurricane Sandy's storm track with the call:

@@

Similarly, you can map binary county exposure based on rain using `map_rain_exposure`. In the case of rain, you must define exposure thresholds for both rainfall and distance. This is because the rain data is observed data, so neglecting to limit "exposed" counties to those close to the storm track may pick up counties far from the storm that experienced a lot of rain over the same period. For example, you can use the following call to map rain exposure to Tropical Storm Allison (2001), where a county is defined as exposed to the storm if cumulative rainfall from five days before to three days after the storm's closest approach was 175 mm or more and the county was within 500 km of the storm's track:

@@

You can use the `map_wind_exposure` function to map counties exposed to a certain winds during a storm. For example, to identify all counties exposed to sustained winds of 20 m / s or higher during Hurricane Katrina, you can run:

@@

Event listings from the NOAA Storm Events database were linked to storms based on the start date of the event being within a five-day window of the date the storm was closest to the county and the storm coming within 500 km of the county. The [`noaastormevents` package](https://github.com/zailchen/noaastormevents) was used to create this dataset. For more information on the data, see the help file for the `storm_events` dataset (`?storm_events`) or the vignette for the `noaastormevents` package.

@@

If you would like to add a storm's track to one of the maps of county-level exposure, you can do so using the `map_tracks` function. This function can be used either to map a storm track on a blank U.S. map or to add a track to an existing map, if the map is a `ggplot` object (as are all maps generated in the previous section).

@@

As another example, to map all tracks for storms in 2005, you can run:

```{r fig.width = 5.5, fig.height = 4}
library(dplyr)
library(tidyr)
storms_2005 <- hurr_tracks %>%
        select(storm_id) %>%
        distinct() %>%
        separate(storm_id, c("name", "year"), sep = "-", remove = FALSE) %>%
        filter(year == "2005")
map_tracks(storms = storms_2005$storm_id) 
```

The `map_tracks` function is used in all the mapping functions described in the previous section, to add the storm's track to the exposure maps. However, if you would like to customize the appearance of the storm's track on the map, you can do so by plotting the exposure map without the storm track (`add_track = FALSE`) and then adding the track yourself with the `map_tracks` function. For example, to change the color of the storm track in a map of flood event exposure during Hurricane Floyd (1999) and to add points showing the available observations for the storm in the Best Tracks hurricane tracking data, you can run:

@@

The `hurricaneexposure` package also has several functions that can input a list of counties and output a list of all of the storms to which each county was exposed over a certain period, based on one of the hazard metrics. These exposure listings include the storm dates, making them easy to integrate into time series datasets of health or other impact data.

The `county_rain` function takes a list of county FIPS codes, bounds on the starting and ending years of the analysis, and thresholds to define rain-based exposure (cumulative rainfall and distance from the storm's track, as explained for the binary rain maps shown in an earlier section) and creates a list of all storms that met these thesholds for the counties. For example, to get a dataset of all the storms to which Orleans Parish (FIPS 22071), and Newport News, Virginia (FIPS 51700), were exposed between 1995 and 2005, where "exposed" means that the storm passed within 100 kilometers of the county center and the rainfall over a three-day window of the date of closest approach was 100 millimeters or more, you can run:

@@

The returned dataframe includes the estimated maximum values during the storm for sustained winds (`vmax_sust`) and gust winds (`vmax_gust`), both in m / s, as well as the duration for which sustained and gust winds were over 20 m / s in the county (`sust_dur` and `gust_dur`, respectively). It also includes some of the variables on the date and time of closest approach and the distance at the storm's closest approach.

The `county_wind` function allows some further specifications when identifying the storms to which a county was exposed based on wind estimates. First, it allows the use of alternative wind variables. While the default is to use the maximum sustained wind speed experienced in the county during the storm, it is also possible to base the exposure metric on the maximum gust winds (`wind_var = "vmax_gust"`) or on the duration of either sustained or gust winds at or above 20 m / s (`wind_var = "sust_dur"` and `wind_var = "gust_dur"`, respectively). For example, to determine which storms brought Orleans Parish sustained winds of 34 knots or higher (17.5 m / s or higher) for an hour or more (`wind_limit = 60`; for wind durations, this wind limit should be specified in minutes), you can run:

```{r}
county_wind(counties = "12086", start_year = 1988, end_year = 2015, 
            wind_var = "sust_dur", wind_limit = 60)
```

Further, the `county_wind` function allows you to pull wind estimates from an alternate source. By default, this function uses wind estimates for each county from a wind model (see the vignettes for the `stormwindmodel` package for much more detail on this modeling process). However, you can also pull estimates based on the wind radii from the Extended Best Tracks dataset by using the option `wind_source = "ext_tracks"`. See the help file for the `ext_tracks_wind` dataset that comes with `hurricaneexposuredata` for more details on this data source and how the county-specific exposures were determined from the data source for use in the `county_wind` function. It is important to note that this data results in wind estimates with breaks at 34 knots, 50 knots, and 64 knots, rather than continuous estimates. This means that a sustained wind estimate from this data source of 17.4896 m / s (34 knots) is estimating that the county had maximum wind speeds of 34 knots or higher during the storm, but not as high as 50 knots. Further, the `dur_sust` value when using this data source is based on number of minutes with winds at or above 34 knots (rather than the 20 m / s value used for durations for the modeled wind data). Gust durations can not be determined using this Extended Best Tracks dataset.

The Extended Best Tracks wind radii dataset is more tied to observations during a specific storm than the modeled data. In certain cases, especially storms in extratropical transition, this dataset might be preferable to the modeled wind speeds, even though it provides less continuous estimates. This source of wind data can be specified using the `wind_source` option in the `county_wind` function. For example, to generate a list of all storms in Orleans Parish with maximum sustained winds of 34 knots or more based on the Extended Best Tracks wind radii, you can run:

```{r}
county_wind(counties = "12086", start_year = 1988, end_year = 2015, 
            wind_var = "vmax_sust", wind_limit = 17.4, wind_source = "ext_tracks")
```

@@

Similar functions exist for wind exposure (`multi_county_wind`) and distance exposure (`multi_county_distance`). In some cases, these multi-county functions may have less functionality than their `county_*` counterparts. For example, the `multi_county_wind` function currently only allows use of the sustained wind speeds for the exposure metric, while the `county_wind` function allows the use of sustained winds, gust winds, or durations of either.
